Results

### 2025-10-29 index.qmd
docs: methods (#1)

---
title: Evolution of solar wind discontinuities in the inner heliosphere
subtitle: PSP and Earth conjunctions and alignments
author:
  - name: Zijin Zhang
    email: zijin@ucla.edu
    orcid: 0000-0002-9968-067X
    affiliation: University of California, Los Angeles
    roles: writing
    corresponding: true
  - name: Anton V. Artemyev
    email: aartemyev@igpp.ucla.edu
    orcid: 0000-0001-8823-4474
    affiliation: University of California, Los Angeles
  - name: Vassilis Angelopoulos
    email: vassilis@ucla.edu
    orcid: 0000-0001-7024-1561
    affiliation: University of California, Los Angeles
  - name: Tai Phan
    email: phan@ssl.berkeley.edu
    orcid: 0000-0002-6924-9408
    affiliation: Space Sciences Laboratory, University of California, Berkeley
---

# Motivation

Studying the radial evolution of solar wind discontinuities from synergistic observations of PSP and Earth-orbiting missions (ARTEMIS, Wind) during aligned intervals.

- How does the discontinuities change with the radial distance from the Sun?

- How is solar wind discontinuities formed? What is the physical mechanisms?

  - Generated at or near the sun?

  - Locally generated in the interplanetary space by turbulence?

# Introduction

@lotekarKineticscaleCurrentSheets2022
@perroneCoherentEventsIon2020

The study of the spatial evolution of solar wind discontinuities is important for understanding the dynamics of the solar wind

Previous statistical studies of solar wind discontinuities in the inner heliosphere @marianiStatisticalStudyMagnetohydrodynamic1983 @liuCharacteristicsInterplanetaryDiscontinuities2021

Alignement study of solar wind discontinuities @artemyevDynamicsIntenseCurrents2018

By indentifying the periods of conjunctions and alignments between PSP and Earth-orbiting missions

@hudsonDiscontinuitiesAnisotropicPlasma1970

“Knetter [2005] extended this work to intervals in 2002 and 2003 which also included fast wind intervals. He obtained similar results to above and amassed a final total collection of 366 discontinuities. All of the discontinuities had normal field components consistent with zero. He concluded that the discontinuities are best explained as TDs regardless of the amount of dB/B.” ([Vasquez et al., 2007, p. 2](zotero://select/library/items/VPNY4UFZ)) ([pdf](zotero://open-pdf/library/items/DNN2ADMT?page=2&annotation=9Q8CVTE9))

# Dataset and method

Our study focuses on the evolution of solar wind discontinuities within plasma parcels as they undergo expansion. We analyze datasets from the Parker Solar Probe (PSP) and Earth-orbiting missions (e.g., ARTEMIS, DSCOVR, and Wind) at various radial distances in the inner heliosphere, during periods of conjunctions and alignments between PSP and these Earth-orbiting missions.

To achieve this, we utilize a configuration where two spacecraft are radially aligned in the Heliographic (HG) frame, with the longitude fixed on the Sun. It is important to note that the plasma parcels observed by the two spacecraft may differ in composition, as solar wind particles propagate at varying velocities. For instance, protons often exhibit a beam in their velocity distribution function, while the thermal speed of electrons is much higher than their bulk speed. Nonetheless, observing plasma that passes through both spacecraft offers the best opportunity to study the evolution of solar wind discontinuities with minimal influence from other processes during propagation. It is also expected that these plasma parcels evolve due to heating and other processes.

<!-- ::: {#fig-plasma-adiabatic-evolution .column-page}
![](figures/evolution/plasma-adiabatic-evolution_e2-DSCOVR-next.svg)

![](figures/evolution/plasma-adiabatic-evolution_e2-DSCOVR.svg)

Comparison of plasma dimensionless quantities (plasma beta and normalized plasma speed by Alfven speed) between PSP (red) and DSCOVR (blue). Yellow dots are predicted plasma beta and normalized speed using formulary from @perroneRadialEvolutionSolar2019.
::: -->

The plasma parcels in question are large-scale structures. As the scale size decreases below the mesoscale [@viallMesoscaleStructureSolar2021] (with radial sizes ranging from approximately 5 × 10³ km to 10⁷ km), it becomes increasingly unlikely to identify the same structure, depending on the latitude difference between the two spacecraft. Furthermore, although large-scale structures (especially density enhancements) can be identified as passing through both spacecraft [@berriotIdentificationSinglePlasma2024], it is nearly impossible to detect kinetic-scale structures, such as solar wind discontinuities (with time scales on the order of seconds), by both spacecraft simultaneously. Therefore, our study focuses on the statistical properties of solar wind discontinuities within the "same" large-scale structures that pass through both spacecraft.

After identifying the periods of conjunctions and alignments between PSP and Earth-orbiting missions, we apply the methodology described in [@???] to identify and characterize solar wind discontinuities. Generally, discontinuities observed by PSP are shorter (1-5 seconds) and exhibit higher current density. Consequently, we use different time windows for identifying discontinuities: 15 seconds for PSP and 30 seconds for Earth-orbiting missions. A example of the discontinuities observed by PSP is shown in @fig-examples.

::: {#fig-examples .column-page}

![](figures/examples/psp_e2_example.png)

Solar wind discontinuities examples observed by PSP fitted by a logistic function in LMN coordinates.
:::

<!-- 
PSP:
SWEAP consists of four separate sensors that measure the ions and electrons that make up the plasma in the solar wind and solar corona. The Solar Probe ANalyzer (SPAN) instruments are electrostatic analyzers (ESAs) that use curved plates to measure the 3-D particle distribution from within the shadow cast by the spacecraft thermal shield. Two of the sensors are co-located on the ram side of the spacecraft: SPAN-Ai measures ions, and SPAN-Ae measures electrons. SPAN-B measures electrons and is on the anti-ram side of the spacecraft. The Solar Probe Cup (SPC) is a Faraday Cup instrument that is directly illuminated by the Sun and measures both ions and electrons.

"Looking at your intervals, which occurred during the early phase of the mission when PSP was far from the Sun, I realize that the plasma data wasn't good for your purpose because (1) SPAN-ion during at least the first 3-4 orbits did not measure the solar wind well, since a good portion of the solar wind distributions was out of view of that instrument, and (2) data from SPC (Farady Cup) does not give good temperature data. I am wondering if you have conjunction events that occurred later, say starting Encounter 7, when data from SPAN would be accurate. If not, some parts of Encounters 4-6 might be OK."
-->

For this study

We utilized high-resolution magnetic field and plasma data from the FIELDS and SWEAP instruments aboard the Parker Solar Probe (PSP). For the plasma data presented in the figures of this paper, we used the dataset from SPAN-i, which measures ions in the solar wind. Although ion data is also available from SPC, the SPAN-i data was chosen due to its higher temporal resolution, making it more suitable for analyzing discontinuities. Our analysis confirms that the results obtained from SPC are consistent with those derived from SPAN-i data. It is also worth noting that sqtn data from the Radio Frequency Spectrometer (RFS) and Low-Frequency Receiver (LFR) for density measurements is available. However, we generally find that the results obtained from RFS/LFR do not differ significantly from those derived from SPAN-i data. Due to the limitations of SPAN-i data during the early mission phases, when a significant portion of the solar wind distributions was out of the instrument’s view, our study concentrates on data beginning with Encounter 7, where SPAN-i provides more accurate measurements.

<!-- 
ARTEMIS:
We use the following data collected by ARTEMIS P1: magnetic fields with a ∼4s spin resolution (Auster et al. 2008), the ion bulk velocity v, and the ion (assumed to be proton) temperature (scalar) Tp calculated from velocity distributions measured at spin resolution by the Electrostatic Analyzer (ESA; McFadden et al. 2008). Being designed for measurements of a hot, rarefied plasma in Earth’s magnetosphere, ESA typically overestimates the proton temperature in the solar wind, because the solar wind proton beam is narrower than ESA’s energy and angular resolution.
TODO: check if the ESA is in solar wind mode
-->

We utilized data collected by ARTEMIS, including magnetic fields measured at approximately 5 Hz resolution by the Fluxgate Magnetometer [@austerTHEMISFluxgateMagnetometer2008], as well as plasma velocity ($v$) and density ($n$), which were derived from velocity distribution data obtained by the Electrostatic Analyzer with a resolution of around 4 seconds [@mcfaddenTHEMISESAPlasma2009]. It is important to note that the ESA tends to overestimate proton temperature in the solar wind due to the narrower proton beam compared to the ESA’s energy and angular resolution. Therefore, any findings related to temperature anisotropy should be interpreted with caution.

<!-- 
WIND:
Wind 3dp, PESA LOW 1 spin resolution ion (proton and alpha) moments
Wind 3D Plasma Analyzer
Instrument Description
The Wind 3DP instrument consists of six different sensors. There are two electron (EESA) and two ion (PESA) electrostatic analyzers with different geometrical factors and field-of-views covering the energy range from 3 eV to 30 keV. There are also a pair of solid state telescopes (SST) that measure electrons with energies up to 400 keV and protons with energie
-->

The following data collected by Wind are used: magnetic fields with a $\sim 11$ Hz resolution measured by the Magnetic Field Investigation (MFI) [@leppingWINDMagneticField1995], the ion (proton and alpha) moments measured by ion (PESA) electrostatic analyzers of the 3D Plasma Analyzer (3DP) [@linThreedimensionalPlasmaEnergetic1995].

> We will also present plasma density estimates delivered by the quasi-thermal noise spectroscopy available at about 7 s cadence (Moncuquet et al. 2020).
> The procedure of CS selection and methodology, briefly described below, is equivalent to those used in the paper

@telloniSpacecraftRadialAlignments2023

<!-- sensing solar wind emitted from the same region on the solar surface, with the caveat that dynamical processes at the chromosphere–corona interface have not dramatically altered the solar footpoint within the time corresponding to the radial propagation from the measurement of one spacecraft to that of the other. -->

# Overview and Example of Current Sheets

> @lotekarKineticscaleCurrentSheets2022 Figure 1 presents an overview of the considered 10 day interval. Over this interval, the spacecraft was at radial distances from 0.17 to 0.24 au, that is, from 35 to 50 solar radii, and remained within the inward magnetic field sector without crossing the heliospheric current sheet (Phan et al. 2020; Szabo et al. 2020). No coronal mass ejections were observed in the considered interval (Phan et al. 2020; Szabo et al. 2020). The spacecraft was approximately corotating with the Sun and connected to the same coronal hole (Bale et al. 2019; Badman et al. 2020). Panels (a)–(d) present 1 minute averages of the magnetic field magnitude, proton flow velocity, densities, and temperatures of protons and electrons as well as plasma density estimates provided by the quasi-thermal noise spectroscopy. The magnetic field magnitude was about 100 nT at the perihelion on November 6, and around 50 nT as the spacecraft moved to a radial distance of 50 solar radii from the Sun on November 1 and 10. The solar wind was typically slow  with a proton flow velocity below 400 km s−1, except for the ≈20 hr period of fast solar wind between November 9 and 10. The three plasma density estimates in panel (c) are consistent with each other within a few tens of percent over the entire interval, except for the fast solar wind interval, where the proton densities were about 4 times smaller than the plasma density estimates provided by the quasi-thermal noise spectroscopy. We will use the electron density estimates, calibrated to best match proton densities and the results of quasi-thermal noise spectroscopy, as plasma density estimates. The proton temperature during the considered interval varied between about 10 and 100 eV, while the electron temperature remained around 30 eV. Panel (e) presents 1 minute averages of proton and electron betas and shows that both quantities varied in the range from 0.1 to 5.
> Figure 2 presents several CSs from our data set.

## Examples of discontinuities

![](figures/examples/psp_e2_example.png)

In general discontinuities observed by PSP are with

- shorter duration (1~5 secs)

  compared with 5-10 secs

- large current density (100-1000 nA/m^2)

  compared with 1-10 nA/m^2

# Results

![](figures/overview-7-all.pdf){#fig-overview}

![](figures/properties_hist-mva.pdf){#fig-properties-hist}

![](figures/joint_properties-mva.pdf){#fig-joint-properties}

![](figures/Alfvenicities.pdf)

![](figures/Q_sonnerup_joint_dist_den.pdf)

![](figures/B_n_ω_subset=true.pdf)

Comparison between Alfven speed and plasma speed change

For rotational discontinuities, the plasma velocity jump across them: the plasma flow velocity $v_l$ (the solar wind velocity projected onto $l$) changes due to changes in $B_l$, $Δv_l=±Δv_A$ with $v_A= B_l / \sqrt{4 n m}$.

![](figures/enc7/dvl.png)

K2: $Δv_l$,l demonstrates a strong correlation with $Δv_{A}$, albeit being consistently smaller.

## The effect of alignment on discontinuity properties

![](figures/thickness_current.png)

![](figures/B_angle.png)

# Discussion

K1: Normalized discontinuities properties does not vary much with radial distance

Thickness and current density relationship, scale-dependent properties of current sheets in the solar wind

# Conclusion

- The properties of the discontinuities is related to the local plasma parameters
  - Thickness => ion inertial length
  - Current density => Alfven velocity (current density)
- Normalized thickness and current density of discontinuities remain constant with radial distance
- Anisotropy of the plasma is expected to be larger near the sun to explain the observed speed change ratio $Δv_i/Δv_A$.

- Normalized thickness and current density of discontinuities remain constant with radial distance (negligible change compared to their spread). In contrast to @artemyevDynamicsIntenseCurrents2018 results that show a decrease in normalized thickness and increase in normalized current density with radial distance.

- Better alignment period has a slightly better agreement of the properties of the discontinuities (normalized thickness, current density, $|\Delta \mathbf{B}/B|$ and rotation angle) between PSP and Earth-orbiting missions.
- $B_N/B$ and in-plane rotation angle are significantly different between PSP and Earth-orbiting missions independent of the alignment period.

# References

-   @velliUnderstandingOriginsHeliosphere2020

-   @artemyevDynamicsIntenseCurrents2018 examined the evolution of transient currents in the solar wind using a data set compiled from observations of the same solar wind ﬂow at Earth’s and Mars's orbits.

    ```         
    > we show that it consists of several processes: discontinuity thinning (decrease in thickness normalized by the ion inertial length), intensification of currents normalized to the proton thermal current (i.e., the product of proton charge, density, and thermal velocity), and increase in the compressional component of magnetic ﬁeld variations across discontinuities.
    ```

# Appendix

>>>>

See [article.qmd](article.qmd) for the main article.

See [notebooks/00_index.qmd](notebooks/00_index.qmd), [notebooks/02_stats.qmd](notebooks/02_stats.qmd), and [notebooks/10_results.qmd](notebooks/10_results.qmd) for the data analysis.

See [github](https://github.com/beforerr/psp_conjunction) for the code.

### 2025-10-29 notebooks/00_index.qmd
docs: methods (#1)

```{julia}
using Dates
using Speasy
using DrWatson
using PSPAnisotropy
using PSPAnisotropy: Wind, PSP, THEMIS
using Discontinuity
using Discontinuity: compute_params!, filter_low_mva_quality
using Discontinuity: compute_Alfvenicity_params!, Alfven_velocity
using SPEDAS
using DimensionalData
using DataFrames, DataFramesMeta
using SpacePhysicsMakie, GLMakie
using SPEDAS: setmeta, setmeta!
using Beforerr: _theme_legend, easy_save, hideylabel!
using PartialFunctions

@@

using Unitful
Unitful.preferunits(u"nA", u"km")
unitify(u) = (*)$u
unitify(u::String) = (*)$uparse(u)
includet("../src/plot.jl")
```

@@

```{julia}
let add_fit = true, add_Va = true
    f = Figure(; size=(450, 400))
    faxs, = plot_candidate(f[1, 1], filter_low_mva_quality(psp_df)[20, :], psp_conf, Second(4); add_fit, legend=nothing, add_Va)
    faxs, = plot_candidate(f[1, 2], filter_low_mva_quality(thm_df)[12, :], thm_conf, Second(2); add_fit, add_Va)
    hideylabel!.(faxs.axes)
    add_labels!(f[1, 1]; position=Left(), padding=(30, -30, 0, 0), labels=1:3, open="(a.")
    add_labels!(f[1, 2]; position=Left(), padding=(30, -30, 0, 0), labels=1:3, open="(b.")
    # faxs, = plot_candidate(f[1, 3], filter_low_mva_quality(wind_df)[2, :], wind_conf, Second(6); add_fit)
    # hideylabel!.(faxs.axes)
    colgap!(f.layout, 1)
    resize_to_layout!(f)
    easy_save("event-example")
    f
end
```

@@

<!-- TODO: add Alfvénicity over time -->

@@

@@

```{julia}
tvars_themis = [
    set(tnorm_combine ∘ THEMIS.B_GSE; ylabel=𝒀.B, labels=𝑳.B_XYZ),
    set(THEMIS.n; ylabel=𝒀.n),
    replace_outliers! ∘ set(THEMIS.V_GSE; ylabel=𝒀.V, labels=𝑳.V_XYZ),
    set(THEMIS.eTemp; ylabel=𝒀.T)
]
```

@@

```{julia}
update_theme!(; Lines=(linewidth=1, alpha=0.9), figure_padding=(1, 1, 1, 15))

for enc in 7:9
    (tmin_psp, tmax_psp), (tmin_wind, tmax_wind) = get_timerange(enc)
    legend = (; position=Top())
    labels = '1':'9'
    padding = (0, 30, 0, -30)
    f = Figure(; size=(1000, 800))
    faxs = plot_PSP(f[1, 1], tmin_psp, tmax_psp, get_vl_ratio_ts(psp_df)...; legend)
    modify_faxs_base!(faxs)
    faxs = plot_Wind(f[1, 2], tmin_wind, tmax_wind - Day(1), get_vl_ratio_ts(wind_df)...; legend)
    modify_faxs_base!(faxs)

    add_labels!(f[1, 1]; labels, open="(a.", padding)
    add_labels!(f[1, 2]; labels, open="(b.", padding)
    display(f)
    easy_save("overview-$enc-all")
end
```

>>>>

```{julia}
using Dates
using Speasy
using DrWatson
using PSPAnisotropy
using PSPAnisotropy: Wind, PSP, THEMIS
using Discontinuity
using Discontinuity: compute_params!, filter_low_mva_quality
using Discontinuity: compute_Alfvenicity_params!, Alfven_velocity
using SPEDAS
using DimensionalData
using DataFrames, DataFramesMeta
using SpacePhysicsMakie, GLMakie
using SPEDAS: setmeta, setmeta!
using Beforerr: _theme_legend, easy_save, hideylabel!, align_ylabels!
using PartialFunctions

@@

using Unitful
Unitful.preferunits(u"nA", u"km")
unitify(u) = (*)$u
unitify(u::String) = (*)$uparse(u)
includet("../src/plot.jl")

@@

```{julia}
let add_fit = true, add_Va = true
    f = Figure(; size=(600, 400))
    psp_row = filter_low_mva_quality(psp_df)[20, :]
    faxs, = plot_candidate(f[1, 1], psp_row, psp_conf, Second(6); add_fit, legend=nothing, add_Va)
    align_ylabels!(faxs.axes)
    thm_row = filter_low_mva_quality(thm_df)[18, :]
    faxs, = plot_candidate(f[1, 2], thm_row, thm_conf, Second(8); add_fit, add_Va)
    hideylabel!.(faxs.axes)
    add_labels!(f[1, 1]; position=Left(), padding=(30, -30, 0, 0), labels=1:3, open="(a.")
    add_labels!(f[1, 2]; position=Left(), padding=(30, -30, 0, 0), labels=1:3, open="(b.")
    # faxs, = plot_candidate(f[1, 3], filter_low_mva_quality(wind_df)[2, :], wind_conf, Second(6); add_fit)
    # hideylabel!.(faxs.axes)

    @info "PSP" psp_row.ΔV_ratio psp_row.ΔV_cosθ
    @info "THEMIS" thm_row.ΔV_ratio thm_row.ΔV_cosθ

    colgap!(f.layout, 1)
    resize_to_layout!(f)
    easy_save("event-example")
    f
end
```

![](../figures/event-example.png)

┌ Info: PSP
│   psp_row.ΔV_ratio = 0.31703819329006777
└   psp_row.ΔV_cosθ = -0.8208860063945036
┌ Info: THEMIS
│   thm_row.ΔV_ratio = 0.3268322029115623
└   thm_row.ΔV_cosθ = -0.9457794099717979

@@

@@

```

@@

@@

```{julia}
using Beforerr: add_labels!, align_ylabels!
update_theme!(; Lines=(linewidth=1, alpha=0.9), figure_padding=(1, 1, 1, 15))

function add_wind_twin!(gl, wind_axs)
    # Create a twin axis for Wind magnetic field and density (multiply by the radial ratio)
    # 1AU / 1Rsun =215
    B_ax = wind_axs.axes[1]
    B_ax2 = Axis(gl[1, 1]; yaxisposition=:right, rightspinevisible=true, rightspinecolor=:darkgray)
    hidexdecorations!(B_ax2, grid=false)
    ylims!(B_ax, 0.1, 9.9)
    ylims!(B_ax2, 0.1 * (0.1)^(-1.59), 9.9 * (0.1)^(-1.59))
    n_ax = wind_axs.axes[2]
    n_ax2 = Axis(gl[2, 1]; yaxisposition=:right, rightspinevisible=true, rightspinecolor=:darkgray)
    hidexdecorations!(n_ax2, grid=false)
    ylims!(n_ax, 3, 16)
    ylims!(n_ax2, 3 * (0.1)^(-1.96), 16 * (0.1)^(-1.96))222
end

function overview_plot(enc)
    (tmin_psp, tmax_psp), (tmin_wind, tmax_wind) = get_timerange(enc)
    legend = (; position=Top())

    f = Figure(; size=(1000, 800))
    psp_axs = plot_PSP(f[1, 1], tmin_psp, tmax_psp, get_vl_ratio_ts(psp_df)...; legend)
    modify_faxs_base!(psp_axs)
    gl = GridLayout(f[1, 2])
    wind_axs = plot_Wind(f[1, 2], tmin_wind, tmax_wind - Day(1), get_vl_ratio_ts(wind_df)...; legend)
    modify_faxs_base!(wind_axs)

    labels = (; labels='1':'9', position=Left(), padding=(32, -32, 0, 0))
    add_labels!(f[1, 1]; labels..., open="(a.")
    add_labels!(f[1, 2]; labels..., open="(b.")
    colgap!(f.layout, 2)

    hideylabel!.(wind_axs.axes)
    add_wind_twin!(gl, wind_axs)
    align_ylabels!(psp_axs.axes)
    easy_save("overview-$enc-all")
    return f
end

overview_plot(7)
# overview_plot.(7:9)
```

![](../figures/overview-7-all.png)
![](../figures/overview-8-all.png)
![](../figures/overview-9-all.png)
